---
title: "Final Project EDA Notebook"
output: html_notebook
---

```{r}
library(dplyr)
library(tidyverse)
library(broom)
library(readxl)
library(modelr)
library(rsample)
```
NOTE: inf_adj_perchild is in $1000.

```{r}
kids <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-09-15/kids.csv')
kids
```
list of vars: 
addCC
CTC -- Child Tax Credit
edservs -- Public spending on education special services by state and year, in $1,000s.
edsubs -- Public spending on education subsidies, including tuition and scholarships, by state and year, in $1,000s.
fedEITC
fedSSI
HCD -- housing and community developments
HeadStartPriv
highered -- higher education expenditures
lib -- libraries
Medicaid_CHIP
other_health
othercashserv
parkrec -- parks and rec
pell -- pell grant
PK12ed -- K12 expenditures
pubhealth
SNAP
socsec -- Public spending on Social Security payments that go to children by state and year, in $1,000s
stateEITC
TANFbasic
unemp -- unemployment
wcomp -- workers comp
```{r}
kids %>%
  group_by(state, variable) %>%
  filter(variable == c("highered","HeadStartPriv", "PK12ed", "SNAP", "HCD", "socsec")) %>%
  summarize(spent_pc = sum(inf_adj_perchild)) %>%
  ggplot() +
  geom_col(aes(state, spent_pc, fill = variable), position = "stack") +
  theme(axis.text.x = element_text(angle = 90)) +
  ggtitle("Money spent by each state per category")
 

```
```{r}
kids %>%
filter(state == "Texas" | state == "Delaware" | state == "Hawaii" | state == "Alaska") %>%
ggplot(aes(year, inf_adj_perchild, color = variable)) +
geom_line() +
facet_grid(cols = vars(state)) +
ggtitle("Money spent over time in each category") +
theme(axis.text.x = element_text(angle = 90)) 
  
```
```{r}
kids %>%
  group_by(variable) %>%
  summarize(sum = sum(inf_adj_perchild)) %>%
  ggplot(aes(variable, sum)) +
  geom_col() +
  theme(axis.text.x = element_text(angle = 90)) + 
  ggtitle("Total spent in US on kids in each category")
```
```{r}
df_income <- read_csv('https://raw.githubusercontent.com/rbyirdaw/US-median-household-income/master/median-household-income-by-state-master.csv')
df_income <-
  df_income %>%
  select("year", "state", "median income")

df_combo <- 
  df_income %>%
  left_join(kids, by = c("year","state")) %>%
  select("year", "state", "median income", "inf_adj_perchild", "variable") %>%
  filter (state!= "United States")

df_combo
names(df_combo)[names(df_combo) == "median income"] <- "median_income"
```
```{r}
df_combo %>%
  group_by(state) %>%
  filter(variable == "PK12ed") %>%
  ggplot(aes(x = median_income, y = inf_adj_perchild, color = state)) +
  geom_point()
```
```{r}
df_combo %>%
  filter(year == "2014") %>%
  select("state", "median_income", "inf_adj_perchild", "variable") %>%
  group_by(state) %>%
  summarize(sum = sum(inf_adj_perchild), income = mean(median_income)) %>%
  ggplot(aes(sum, income, color = state)) +
  geom_point() +
  ggtitle("Total money spent in 2014 on children in comparison to income median that year")
```

```{r}
combined_money <-
  df_combo %>%
  pivot_longer(names_to = "money_type",
               values_to = "amount",
               cols = c(median_income, inf_adj_perchild)) 
```
```{r}
df_combo %>%
  ggplot(aes(state, median_income)) +
  geom_col() +
  theme(axis.text.x = element_text(angle = 90)) 
```
```{r}
library(usmap)
library(ggplot2)
map_df <-
kids %>%
  group_by(variable) %>%
  filter(year == "2014") %>%
  summarize(sum = sum(inf_adj_perchild), state = state)
map_df_2013 <-
 kids %>%
  group_by(variable) %>%
  filter(year == "2013") %>%
  summarize(sum = sum(inf_adj_perchild), state = state) 
```
```{r}
expen_df <- read_csv('2014_asfin_state_totals.csv')
expen_df <- data.frame(lapply(expen_df, as.character), stringsAsFactors=FALSE)

expen_df_2013<-read_csv('2013_asfin_state_totals.csv')
expen_df_2013 <-data.frame(lapply(expen_df_2013, as.character), stringsAsFactors = FALSE)

expen_df <-
  expen_df %>%
  rename(
    "New Hampshire" = New.Hampshire,
    "New Jersey" = New.Jersey,
    "New York" = New.York,
    "New Mexico" = New.Mexico,
    "North Dakota" = North.Dakota,
    "North Carolina" = North.Carolina,
    "Rhode Island" = Rhode.Island,
    "South Carolina" = South.Carolina,
    "South Dakota" = South.Dakota,
    "West Virginia" = West.Virginia
  )
  
expen_df


expen_df_2013 <-
  expen_df_2013 %>%
  rename(
    "New Hampshire" = New.Hampshire,
    "New Jersey" = New.Jersey,
    "New York" = New.York,
    "New Mexico" = New.Mexico,
    "North Dakota" = North.Dakota,
    "North Carolina" = North.Carolina,
    "Rhode Island" = Rhode.Island,
    "South Carolina" = South.Carolina,
    "South Dakota" = South.Dakota,
    "West Virginia" = West.Virginia
  )

```
```{r}
expen_df <-
  expen_df %>%
  pivot_longer(
    names_to = "state",
    cols = everything(),
    values_to = "Total_Expenditures"
  ) %>%
  slice(-c(1)) %>%
  left_join(map_df, by = c("state")) %>%
  group_by(state)%>%
  summarize(total = sum(sum), state=state, Total_Expenditures_2014=Total_Expenditures) 
expen_df$Total_Expenditures_2014 <- as.numeric(expen_df$Total_Expenditures_2014)
expen_df <-
  expen_df %>%
  mutate(norm=(total/Total_Expenditures_2014)) 
expen_df


expen_df_2013 <-
  expen_df_2013 %>%
  pivot_longer(
    names_to = "state",
    cols = everything(),
    values_to = "Total_Expenditures"
  ) %>%
  slice(-c(1)) %>%
  left_join(map_df_2013, by = c("state")) %>%
  group_by(state)%>%
  summarize(total_2013 = sum(sum), state=state, Total_Expenditures_2013=Total_Expenditures) 
expen_df_2013$Total_Expenditures_2013 <- as.numeric(expen_df_2013$Total_Expenditures_2013)
expen_df_2013 <-
  expen_df_2013 %>%
  mutate(norm_2013=(total_2013/Total_Expenditures_2013)) 
expen_df_2013
```


```{r}
plot_usmap(data = expen_df, values = "norm", color = "red") + 
  scale_fill_continuous(
    low = "white", high = "red", name = "Normalized Money Spent on Each", label = scales::comma
  ) + theme(legend.position = "right") + 
  ggtitle("Money Spent per child as a Percentage of Total Expenditures in 2014")


plot_usmap(data = expen_df_2013, values = "norm_2013", color = "red") + 
  scale_fill_continuous(
    low = "white", high = "red", name = "Normalized Money Spent on Each", label = scales::comma
  ) + theme(legend.position = "right") + 
  ggtitle("Money Spent per child as a Percentage of Total Expenditures in 2013")
```
```{r}
plot_usmap(data = map_df, values = "sum", color = "red") + 
  scale_fill_continuous(
    low = "white", high = "red", name = "Money Spent", label = scales::comma
  ) + theme(legend.position = "right") + ggtitle("Money Spent per child in each state, unnormalized")
plot_usmap(data = map_df_2013, values = "sum", color = "red") + 
  scale_fill_continuous(
    low = "white", high = "red", name = "Money Spent", label = scales::comma
  ) + theme(legend.position = "right") + ggtitle("Money Spent per child in each state, unnormalized")
```


