---
title: "Final Project EDA Notebook"
output: html_notebook
---

```{r}
library(dplyr)
library(tidyverse)
library(broom)
```


```{r}
kids <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-09-15/kids.csv')
kids
```
list of vars: 
addCC
CTC -- Child Tax Credit
edservs -- Public spending on education special services by state and year, in $1,000s.
edsubs -- Public spending on education subsidies, including tuition and scholarships, by state and year, in $1,000s.
fedEITC
fedSSI
HCD -- housing and community developments
HeadStartPriv
highered -- higher education expenditures
lib -- libraries
Medicaid_CHIP
other_health
othercashserv
parkrec -- parks and rec
pell -- pell grant
PK12ed -- K12 expenditures
pubhealth
SNAP
socsec -- Public spending on Social Security payments that go to children by state and year, in $1,000s
stateEITC
TANFbasic
unemp -- unemployment
wcomp -- workers comp
```{r}
kids %>%
  group_by(state, variable) %>%
  filter(variable == c("highered","HeadStartPriv", "PK12ed", "SNAP", "HCD", "socsec")) %>%
  summarize(spent_pc = sum(inf_adj_perchild)) %>%
  ggplot() +
  geom_col(aes(state, spent_pc, fill = variable), position = "stack") +
  theme(axis.text.x = element_text(angle = 90)) +
  ggtitle("Money spent by each state per category")
 

```
```{r}
kids %>%
filter(state == "Texas" | state == "Delaware" | state == "Hawaii" | state == "Alaska") %>%
ggplot(aes(year, inf_adj_perchild, color = variable)) +
geom_line() +
facet_grid(cols = vars(state)) +
ggtitle("Money spent over time in each category") +
theme(axis.text.x = element_text(angle = 90)) 
  
```
```{r}
kids %>%
  group_by(variable) %>%
  summarize(sum = sum(inf_adj_perchild)) %>%
  ggplot(aes(variable, sum)) +
  geom_col() +
  theme(axis.text.x = element_text(angle = 90)) + 
  ggtitle("Total spent in US on kids in each category")
```

